How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-23 docs/PhysicsPaper/StokesErosion.tex
NM

% Original ideas
%We numerically simulate the erosion of a porous medium due to an internally flowing fluid, with solid constituents of the medium eroding under the action of surface shear stress. 
%We numerically simulate the erosion of a porous medium due to an internally flowing fluid.  The solid constituents of the porous medium erode under the action of surface shear stress. As the particles disintegrate, they elongate in the direction of the flow, giving rise to anisotropic conductivity of the porous medium.

% dematerialize, disintegrate
% internal, permeating, intervening
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

	Flow-induced erosion acts across a range of scales in the natural world, from massive geological structures sculpted by wind or water \cite{abrams2009growth, perkins2015amplification, mac2020ultra, sharma2022alcove, mac2022morphological}, to mesoscopic patterns formed by surface or internal flows \cite{berhanu2012shape, bertagni2021hydrodynamic}, and down to the granular constituents of porous media as they slowly disintegrate in groundwater flows \cite{zareei2022temporal}. The associated nonlinear feedback between changing shapes and the surrounding flows can imprint across all of these scales, affecting large-scale features as well as small-scale ones, such as the microstructure of porous materials. Porous media encountered in nature, for example, typically exhibit material {\em anisotropy} in that they permit seepage flow in certain directions more easily than in others. Typical materials are more permeable to flow in the longitudinal (or horizontal) direction than transversely by a factor of 5-20 \cite{bear1988dynamics, anderson2015applied}. Most commonly, this material anisotropy is attributed to the sedimentation process, in which, due to the fluid-structure interaction, non-spherical particles tend to settle with their long axis parallel to the plan of deposition \cite{bear1988dynamics}. Controlled experiments, however, have not been performed to test this hypothesis, and other mechanisms may be at work. Here, we use highly-accurate numerical simulations to examine an alternative, and possibly complementary, mechanism: namely, that the flow-induced erosion of the medium's solid constituents contributes to its overall anisotropy.

	Our method merges highly-efficient and highly-accurate boundary-integral equation (BIE) methods with stable interface evolution methods to simulate the erosion of dense suspensions of solid bodies in the Stokes flow regime relevant for groundwater-flow applications \cite{quaife2018boundary}. Originally inspired by related work in the high-Reynolds-number regime \cite{Ristroph2012, Moore2013, Huang2015, MooreCPAM2017}, our method is documented, validated, benchmarked in \cite{quaife2018boundary}, and can simulate the erosion of $O(100)$ solid bodies. The more recent Cauchy reformulation of the BIE and the associated  quadrature formulas allow us to resolve points of near contact between bodies, thus enabling high-fidelity simulation of dense suspensions of erodable bodies \cite{chiu2020viscous}. Though the governing fluid-flow equations (Stokes) are linear, the nonlinear feedback between evolving microstructure and flow give rise to highly complex and anisotropic configurations.

@@

%Maybe cite: \cite{Rycroft2016}
% Cite paper by Ladd "Wormhole formation in dissolving fractures"

@@

The embedded bodies may erode in response to the shear stresses induced by the intervening fluid flow. Erosion typically occurs over much longer timescales than does the fluid flow, permitting a {\em quasi-steady} approximation. In this approximation, the configuration of bodies is held fixed in order to compute the {\em steady} Stokes flow determined by \eqref{eqn:StokesEq}, and then this flow field determines the stresses acting to erode each body. We employ an erosion law in which the local rate of material loss is linearly proportional to the magnitude of the shear stress $\tau$ acting on the surface (CITE previous erosion work). The material loss gives rise to an inward velocity of the solid surface, $\Vn$, pointing in the direction normal to the surface. The erosion law is thus expressed as
\begin{align}
\Vn = \CE \, \abs{\tau}, 
	&\hspace{20pt} \xx \in \gamma, &&\mbox{\em erosion model}, 
\label{Vn0} \\
\tau = -\mu \left( ( \nabla \uu + \nabla \uu^T) \nn \right) \cdot \ss
	&\hspace{20pt} \xx \in \gamma, &&\mbox{\em shear stress}.
\label{tau0}
\end{align}
where $\nn$ is a unit normal vector pointing into each body, $\ss$ is
a unit tangent vector pointing in the counterclockwise direction, and $\CE$ is a material-dependent erosion constant.

@@

Rather than tracking the Cartesian coordinates of each surface, we employ the {\thL} formulation \cite{hou-low-she1994, MooreCPAM2017, mac2022morphological} by tracking the tangent angle $\theta$ as a function of arc length and the total length $L$ of each body.  In this formulation, the curvature-dependent smoothing becomes a linear diffusive term, thus enabling the use of stable, implicit schemes for this stiff term. The remaining nonlinear terms are not stiff and can be treated by explicit time-stepping methods. In particular, we use an exponential integrator for the diffusion term and a Runge-Kutta method for all other terms, both of which are second order in time \cite{quaife2018boundary}.

>>>>

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% dematerialize, disintegrate
% internal, permeating, intervening

@@

	Flow-induced erosion acts across a range of scales in the natural world, from massive geological structures sculpted by wind or water \cite{abrams2009growth, perkins2015amplification, mac2020ultra, sharma2022alcove, mac2022morphological}, to mesoscopic patterns formed by surface or internal flows \cite{berhanu2012shape, bertagni2021hydrodynamic, weady2022anomalous}, and down to granular and porous networks slowly disintegrating in groundwater flows \cite{szymczak2009wormhole, chiu2020viscous, zareei2022temporal}. The associated nonlinear feedback between changing shapes and the surrounding flows can imprint across all of these scales, affecting large-scale features as well as small-scale ones, such as the microstructure of porous materials. Porous media encountered in nature, for example, typically exhibit material {\em anisotropy} in that they permit seepage flow in certain directions more easily than in others. Typical materials are more permeable to flow in the longitudinal (or horizontal) direction than transversely by a factor of 5-20 \cite{bear1988dynamics, anderson2015applied}. Most commonly, this material anisotropy is attributed to the sedimentation process, in which, due to the fluid-structure interaction, non-spherical particles tend to settle with their long axis parallel to the plan of deposition \cite{bear1988dynamics}. Controlled experiments, however, have not been performed to test this hypothesis, and other mechanisms may be at work. Here, we use highly-accurate numerical simulations to examine an alternative, and possibly complementary, mechanism: namely, that the flow-induced erosion of the medium's solid constituents contributes to its overall anisotropy.

	Our method merges highly-efficient and highly-accurate boundary-integral equation (BIE) methods \cite{baker1986boundary, moore2007evaluation} with stable interface evolution methods \cite{hou-low-she1994, Moore2013} to simulate the erosion of dense suspensions of solid bodies in the Stokes flow regime relevant for groundwater-flow applications \cite{quaife2018boundary, chiu2020viscous}. Originally inspired by related work in the high-Reynolds-number regime \cite{Ristroph2012, Moore2013, Huang2015, MooreCPAM2017}, our method is documented, validated, benchmarked in \cite{quaife2018boundary}, and can simulate the erosion of $O(100)$ solid bodies. The more recent Cauchy reformulation of the BIE and the associated  quadrature formulas allow us to resolve points of near contact between bodies, thus enabling high-fidelity simulation of dense suspensions of erodable bodies \cite{chiu2020viscous}. Though the governing fluid-flow equations (Stokes) are linear, the nonlinear feedback between evolving microstructure and flow give rise to highly complex and anisotropic configurations.

@@

%Maybe cite: \cite{Rycroft2016}

@@

The embedded bodies may erode in response to the shear stresses induced by the intervening fluid flow. Erosion typically occurs over much longer timescales than does the fluid flow, permitting a {\em quasi-steady} approximation. In this approximation, the configuration of bodies is held fixed in order to compute the {\em steady} Stokes flow determined by \eqref{eqn:StokesEq}, and then this flow field determines the stresses acting to erode each body. We employ an erosion law in which the local rate of material loss is linearly proportional to the magnitude of the shear stress $\tau$ acting on the surface \cite{Ristroph2012, Moore2013, Mitchell2016, MooreCPAM2017, HewettJFS2017, quaife2018boundary, chiu2020viscous}. The material loss gives rise to an inward velocity of the solid surface, $\Vn$, pointing in the direction normal to the surface. The erosion law is thus expressed as
\begin{align}
\Vn = \CE \, \abs{\tau}, 
	&\hspace{20pt} \xx \in \gamma, &&\mbox{\em erosion model}, 
\label{Vn0} \\
\tau = -\mu \left( ( \nabla \uu + \nabla \uu^T) \nn \right) \cdot \ss
	&\hspace{20pt} \xx \in \gamma, &&\mbox{\em shear stress}.
\label{tau0}
\end{align}
where $\nn$ is a unit normal vector pointing into each body, $\ss$ is
a unit tangent vector pointing in the counterclockwise direction, and $\CE$ is a material-dependent erosion constant.

@@

Rather than tracking the Cartesian coordinates of each surface, we employ the {\thL} formulation \cite{hou-low-she1994, Moore2013, MooreCPAM2017, mac2022morphological} by tracking the tangent angle $\theta$ as a function of arc length and the total length $L$ of each body.  In this formulation, the curvature-dependent smoothing becomes a linear diffusive term, thus enabling the use of stable, implicit schemes for this stiff term. The remaining nonlinear terms are not stiff and can be treated by explicit time-stepping methods. In particular, we use an exponential integrator for the diffusion term and a Runge-Kutta method for all other terms, both of which are second order in time \cite{quaife2018boundary}.
